---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(width = 300)
library(tidyverse)
library(sensemakr)
library(sensewls)
```



# sensewls

Please contact Leonard Wainstein (lwainstein@reed.edu) with questions. Thank you to Wolfgang Brightenburg (wbrigh@uw.edu) for creating the initial versions of this package and its README. 

## Overview

Researchers often seek the causal effect of a treatment, $D$, on an outcome of interest, $Y$. 
In observational settings, estimating this unbiasedly requires accounting for all of the confounders in the relationship between the treatment and the outcome. 
One common estimation approach here is a weighted least squares (WLS) linear regression of the outcome on the treatment and some observed covariates, $X$, where the weighting is meant to reduce dependence on the estimated linear model ([Ho et al., 2007](https://www.cambridge.org/core/journals/political-analysis/article/matching-as-nonparametric-preprocessing-for-reducing-model-dependence-in-parametric-causal-inference/4D7E6D07C9727F5A604E5C9FCCA2DD21)). 
However, even before considering violations to modeling assumptions implied by the linear regression or the weights, a first order concern is the risk of unobserved confounding that leads to persistent biases regardless of the conditioning technology used. The claim that all variables that must be accounted for to achieve unbiasedness is unlikely to hold in most real-world cases. The resulting bias in the estimate will be driven by the extent to which unobserved confounders, $Z$, are related to the treatment and outcome conditionally on the covariates. Transparency thus requires that we assess the sensitivity of one's conclusions to unobserved confounding, i.e., "sensitivity analysis".

The `sensewls` package implements the sensitivity analysis tools
proposed by [Wainstein and Hazlett (2026)](https://www.arxiv.org/abs/2508.02954), in which standardized statistics and covariate bounds are introduced to quantify the
sensitivity of WLS estimates of causal estimands to unobserved
confounding. These tools are the weighted extension to the tools proposed by [Cinelli and Hazlett (2020)](https://academic.oup.com/jrsssb/article/82/1/39/7056023), which are implemented through the [`sensemakr`](https://cran.r-project.org/web/packages/sensemakr/index.html) package. These tools allow researchers to quantify how strong an unobserved
confounder (or group of confounders) would need to be in order to change
research conclusions, with strength measured in either absolute terms
through weighted partial correlation, or in relative terms, measured in
relation to the strength of observed covariate(s).

To demonstrate the general workflow of `sensewls` and its reliance on domain
knowledge, we provide the following tutorial through an extended
example, studying the average effects of exposure to violence in Darfur
on attitudes towards peace. The data is accessed through the `sensemakr` package. The following example is a replication of the analysis done by [Wainstein and Hazlett (2026)](https://www.arxiv.org/abs/2508.02954). Further details and discussion on this setting can be found in [Hazlett (2020)](https://journals.sagepub.com/doi/abs/10.1177/0022002719879217).

## Installation

You can install the development version of `sensewls` from [GitHub](https://github.com/) with:

```{r, echo=TRUE, eval=FALSE}
# install.packages("devtools")
devtools::install_github("lwainstein/sensewls")
```


## Example


The `sensewls` function takes in a variety of arguments. For specifying the variables of interest, the function has the following four
arguments:

-   `df` A dataframe containing `treatment`, `outcome`, and `covars`.

-   `treatment` A string indicating the column name of the treatment
    variable.

-   `outcome` A string indicating the column name of the outcome
    variable.

-   `covars` A character vector of column names of the relevant
    covariates in the dataframe to use in the regression.

To specify the weighting mechanism or weights, we must specify:

-   `w` An argument for the weighting method, which may be (1) a
    numeric vector of weights; (2) a string for one of the preset weighting
    functions; or (3) a custom weighting function. For a
    preset weighting method, three mechanisms are currently implemented.
    A simple implementation of inverse propensity score weights through a
    logistic regression of the treatment on all covariates may be used through setting
    `w` to the string `"prop_score"`. Additionally, the `matchit()`
    function from the [`MatchIt`](https://cran.r-project.org/web/packages/MatchIt/index.html) package and the entropy balancing
    weights function `ebalance()` from the [`ebal`](https://cran.r-project.org/web/packages/ebal/index.html) package may be used by
    setting `w` to the strings `"matchit"` and `"ebal"`, respectively.
    Additional arguments to these methods may be passed at the end of
    the `sensewls` call. Further information on custom weight functions can
    be found in the custom weight function section of this README.

-   `estimand` A string indicating the desired estimand. Only necessary
    if `w` is a preset weighting method.

-   `normalize` A logical indicating whether or not to normalize the weights `w` per [Wainstein and Hazlett (2026)](https://www.arxiv.org/abs/2508.02954) when either providing a numeric vector of weights, or a custom weighting function. Weights are automatically normalized when `w` is a preset weighting method.

To relate an unobserved confounder's strength to that of one or multiple
observed covariates, we need to supply `sensewls` the following arguments:

-   `bounding_covars` A list of character vectors containing covariates
    to bound by.

-   `semiweights` An optional list of semiweight vectors if `w` is a
    numeric weights vector. Providing semiweights is necessary for computing
    covariate bounds, and omitting a weight function to calculate them
    requires to the user to supply them directly. The semiweights list
    should be sorted such that the $i$th entry corresponds to the set
    of semiweights formed when constructing weights while omitting the
    $i$th set of covariates in the `bounding_covars` list.

-   `kd` A numeric vector of considered covariate bound strengths for association with the treatment variable.

-   `ky` A numeric vector of considered covariate bound strengths for association with the outcome variable.


To compute confidence intervals and $RV_{\alpha}$, `sensewls` takes the
following seven arguments:

-   `inference` A string that indicates which inference option is used. The options are: (1) `"boot"`: Non-parametric bootstrap that re-estimates the weights in each bootstrap sample. This is only allowed if a preset weighting function or an appropriate, user-supplied weighting function is chosen for `w`. (2) `"fix_boot"`: Non-parametric bootstrap that fixes the weights, and samples them along with the data in each bootstrap sample. (3) `"HC0"`: (Adjusted) heteroscedastic-consistent standard errors ([White, 1980](https://www.jstor.org/stable/1912934)), without any finite sample corrections. (4) `"HC1"`: (Adjusted) heteroscedastic-consistent standard errors with finite sample correction ([MacKinnon and White, 1985](https://www.sciencedirect.com/science/article/abs/pii/0304407685901587)). (5) `"none"`: No uncertainty estimates are provided. The default is `"HC1"`.

- `ci_method` A string indicating whether to obtain confidence intervals via the normal approximation method (`"normal"`), or the percentile method (`"percentile"`) when bootstrap inference is used (`inference` is `"boot"` or `"fix_boot"`). Default is `"normal"`.

-   `cluster` An optional string indicating a column from the data frame `df` for which to cluster standard errors/bootstrap. If `inference` is `"HC0"` or `"HC1"`, then  cluster-robust standard errors ([White, 1984](https://www.sciencedirect.com/book/monograph/9780127466507/asymptotic-theory-for-econometricians)) are formed.  If `inference` is `"boot"` or `"fix_boot"`, then a cluster bootstrap is performed. Defaults to `NULL`.

-   `alpha` Level of significance used for $RV_{\alpha}$ and the
    confidence intervals. Defaulted to `alpha = 0.05`.

-   `B` Number of bootstrap samples if used. Defaulted to 500.

-   `par` Boolean for whether bootstrapping should be parallelized
    through the `parallel` library. Defaulted to `FALSE`. Only supported
    on MacOS.

-   `ncpus` Number of cpus used if `par` is set to `TRUE`. Defaulted to
    `NULL` and must be supplied if `par` is set to true.

We will now use the `sensewls` function to perform a sensitivity analysis on
a WLS estimate of the ATE using inverse propensity score weights. Note
that the exchangeable unit in this setting is not the individual
respondent, but rather the village that they reside in. Hence, we
utilize a cluster bootstrap as our method of inference, specifying such
through setting the `inference` argument to `"boot"` and providing
the clustering variable `village` as the value for the `cluster` argument.
Omitting an input to `cluster` while utilizing a bootstrap method for inference
defaults to the nonparametric bootstrap where individual rows are
resampled instead of sampling whole groups. We use the percentile method for confidence intervals
after bootstrapping by setting the `ci_method` argument to `"percentile"`.

```{r}
### Read in data
library(sensemakr)
data(darfur)

### Only retain villages in which there are both treated/control
meantreat <- tapply(darfur$directlyharmed, darfur$village, FUN = mean)
meantreat <- meantreat[!is.na(meantreat)]
tvar <- names(meantreat[meantreat>0 & meantreat<1])
tvar <- sort(tvar)
tvar <- data.frame("village" = tvar)
df<- merge(tvar, darfur)

### Run sensitivity analysis 
set.seed(1)
sensewls_darfur <- sensewls(
                    df = df,
                    treatment = "directlyharmed",
                    outcome = "peacefactor", 
                    covars = c("village",
                                "female",
                                "age", 
                                "farmer_dar",
                                "herder_dar",
                                "pastvoted",
                                "hhsize_darfur"), 
                    bounding_covars = list("female" = c("female")), 
                    kd = 1:2,
                    ky = 1:2, 
                    w = "prop_score",
                    semiweights = NULL,
                    estimand = "ATE",
                    inference = "boot", 
                    ci_method = "percentile",
                    cluster = "village",
                    par = F,
                    B = 1000
                  )
```

We can inspect the WLS estimate, standard error, and confidence interval by calling:

```{r}
sensewls_darfur$wls_data
```

This estimation approach indicates that direct harm positively influence attitudes toward peace, and that the estimate is statistically significant at the $\alpha = 0.05$ level.

### Sensitivity Statistics

For the minimal study of the sensitivity of the WLS estimate, the `sensewls` call
returns three key sensitivity statistics to complement the standard
regression outputs of the effect estimate, estimated standard error, and
a confidence interval:

```{r}
sensewls_darfur$RV
```

These sensitivity statistics include: (1) $R^2_w(Y\sim D|X)$, the weighted
partial $R^2$ between the treatment and outcome conditional on
the covariates (`wR2.YD.X`); (2) the robustness value $RV_{q=1}$, or the strength of the unobserved confounding necessary to reduce the
estimated effect to zero (`RV_{q=1}`); and (3) the robustness value $RV_{\alpha}$, or the strength of the unobserved confounding
necessary to reduce the estimated effect as to be statistically
indistinguistable from a null effect at the $\alpha$ significance level (here, `RV_{alpha=0.05}`).
In detail,

(1) $R^2_w(Y\sim D|X)$ serves as the
    minimal amount of weighted residual variance needed to be explained
    by a confounder in order the nullify the observed effect, under the
    extreme scenario in which the confounder is responsible for all
    noise in the regression on the outcome. The observed weighted
    partial correlation $R^2_w(Y\sim D|X) = 0.022$ may be interpreted such
    that, in the extreme scenario in which we assume that some
    unobserved confounders fully explain all of the unaccounted variance
    in the outcome `peacefactor`, these same confounders would need to
    explain at least 2.2% of the variance in the treatment to nullify
    the observed effect.

(2) The robustness value $RV_{q=1}$ quantifies how much the unobserved
    strengths $R^2_{w}(Y\sim Z|X, D)$ and $R^2_{w}(D \sim Z|X)$
    need to be in order to nullify the estimated effect, presuming the
    confounder exerts equal influence on the treatment and outcome
    variables. In this setting, observing that $RV_{q=1} = 0.139%$ means an
    unobserved confounder must explain at least 13.9% of the weighted
    residual variance in both the treatment and outcome to reduce the
    point estimate of the estimated effect to zero.

(3) The robustness value $RV_{\alpha=0.05}$ quantifies how large the strengths $R^2_w (Y\sim Z|X, D)$ and $R^2_w (D \sim Z|X)$ need to be for
    the WLS estimate to fail to be statistically distinguishable from
    zero at the $\alpha=0.05$ significance level (presuming the confounder
    influences the treatment and outcome equally). The observed value
    of $RV_{\alpha=0.05} = 0.058$ posits that unobserved confounders which
    account for 5.8% of the weighted residual variance have enough
    influence to render the estimate statistically
    insignificant.

The above quantities may be used to summarize the strengths necessary
for unaccounted confounders to alter the results of the analysis. Domain
knowledge is necessary to argue whether possible confounders meet these
criteria.

### Covariate Bounds

While the above sensitivity statistics report how strong unobserved confounders would need to be to change our results, they do not tell us if such confounders are plausible. For this, it is useful to consider relative claims on the strengths of
unobserved confounders, such as an unobserved confounder being X times "as strong" as an observed covariate. To do this, we can inspect the covariate bound estimates with:

```{r}
sensewls_darfur$Covariate_Bound_Estimates
```


Our analysis reports an initial estimate that is positive and
significant and the $\alpha = 0.05$ level, suggesting that direct harm
increased attitudes towards peace. Additionally, the sign and
significance of the ATE estimate are robust to a confounder as
strong as `female`, a covariate argued to be influential for
the treatment as discussed in [Hazlett (2020)](https://journals.sagepub.com/doi/abs/10.1177/0022002719879217). If such a confounder were
present, it would have strengths $R_w^2 (D \sim Z | X) = 0.011$ and $R_w^2 (Y \sim Z | X, D) = 0.108$, and the adjusted estimate would remain
positive (+0.069) and statistically significant at the 0.05 level. However, a confounder that is
two times as strong as `female` would render our estimate
statistically insignificant at the $\alpha = 0.05$ threshold, as the adjusted 95\% confidence interval (-0.005, 0.097) contains 0. 

One may also visualize the adjusted estimates from these estimated covariate bounds using the `plot_sensewls()` function:
```{r}
plot_sensewls(sensewls_darfur, xlim = c(0,0.25), ylim = c(0,0.4))
```

### Changing values of $\alpha$ and $q$

When utilizing the bootstrap for inference, it may be computationally
expensive to re-compute analyses to investigate inference at varying
significance levels. If we'd like to view our results when varying $\alpha$ (for $RV_{\alpha}$ and confidence intervals)
or $q$ (for $RV_q$), we can use the `alter_sensewls()` function to adjust our robustness values
and confidence intervals accordingly:

```{r}
### Alter output
new_output <- alter_sensewls(sensewls_darfur, alpha = 0.01, q = 0.9)

### Re-inspect results
new_output$wls_data
new_output$RVs
new_output$Covariate_Bound_Estimates
```

Setting our Type I error rate to a higher standard of $\alpha = 0.01$,
we find that our findings still support a statistically
significant estimate when a confounder as strong as `female` is
present, but just barely. 

### Custom Weight Functions

Beyond utilizing the preset weighting functions, `sensewls` supports users
supplying their own custom weighting function, or vectors of weights and
semiweights.

Custom weight functions require a specific syntax, requiring the
following three arguments: 

- `treatment`, a string indicating the column name of the treatment variable.
- `covars`, a character vector of the column names for the covariates to weight for.
- `df`, a data frame containing the columns named by `treatment` and `covars`.

Additionally, the weight function is expected to only output a vector of
weights ordered such that the $i$th weight corresponds to the $i$th row
in `df`. Below is an example of an implementation of an inverse propensity
score weighting function for the ATE. Additionally, we use the `normalize=TRUE` 
option in the `sensewls` function to normalize the weights per [Wainstein and Hazlett (2026)](https://www.arxiv.org/abs/2508.02954).

```{r}
### Create custom inverse propensity score weighting function for the ATE
custom_ipw <- function(treatment, covars, df){

  psmodel <- glm(as.formula(paste(treatment,
                                  " ~ ", 
                                  paste(covars, collapse = "+"))), 
                 family = binomial(link="logit"),
                 data = df)

  d <- df[, treatment]
  weights <- rep(0, nrow(df))
  weights[d==0] <- 1 / (1 - psmodel$fitted.values[d==0])
  weights[d==1] <- 1 / psmodel$fitted.values[d==1]

  return(weights)
}

### Run sensewls with the custom_ipw function
set.seed(1)
output_customewights <- sensewls(treatment = "directlyharmed",
                  outcome = "peacefactor", 
                  covars = c("village",
                              "female",
                              "age", 
                              "farmer_dar",
                              "herder_dar",
                              "pastvoted",
                              "hhsize_darfur"), 
                  bounding_covars = list("female" = c("female")), 
                  kd = 1:2,
                  ky = 1:2,
                  df = df, 
                  w = custom_ipw,
                  normalize = T,
                  semiweights = NULL,
                  inference = "boot", 
                  ci_method = "percentile",
                  cluster = "village",
                  B = 1000, 
                  par = T, 
                  ncpus = 5)
```

You will notice that the non-inference results are exactly the same as when using the default inverse propensity score weights for the ATE by `w = "prop_score"` and `estimand = "ATE"`. The inference-related results are very close, but have slight differences due to the randomness of the bootstrap. 

```{r}
### Check point estimates
output_customewights$wls_data
sensewls_darfur$wls_data

### Check robustness values
output_customewights$RVs
sensewls_darfur$RVs

### Check covariate bounds
output_customewights$Covariate_Bound_Estimates
sensewls_darfur$Covariate_Bound_Estimates

```




## References

Cinelli, C. and Hazlett, C. (2020). Making sense of sensitivity: Extending omitted variable bias. *Journal of the Royal Statistical Society: Series B (Statistical Methodology), 82*(1):39–67.

Hazlett, C. (2020). Angry or weary? How violence impacts attitudes toward peace among Darfurian refugees. *Journal of Conflict Resolution, 64*(5):844–870.

Ho, D. E., Imai, K., King, G., and Stuart, E. A. (2007). Matching as nonparametric preprocessing for reducing model dependence in parametric causal inference. *Political Analysis, 15*(3):199–236.

MacKinnon, J. G. and White, H. (1985). Some heteroskedasticity-consistent covariance matrix estimators with improved finite sample properties. *Journal of Econometrics*, 29(3):305-325.

Wainstein, L. and Hazlett, C. (2026). Sensitivity of weighted least squares estimators to omitted variables. *arXiv preprint arXiv:2508.02954*.

White, H. (1980). A heteroskedasticity-consistent covariance matrix estimator and a direct test for heteroskedasticity. *Econometrica: Journal of the Econometric Society*: 817-838.

White, H. (1984). Asymptotic theory for econometricians. Academic press. 



